Poseidon Adventure

### 2019-06-13 TestCalcZ2.Rmd
lets call calc_Z good

Nope. We have a constant value punctuated by occassional variation. Seems wrong. Variation doesn't look anything like the txt file either. Perhaps the 5x yr timesteps are messing this up?

Time to step through line by line again. Revised as below. Better (apportions annual recruitment to timesteps), but not sure this is fully correct?

```{r calczsnippet, eval=FALSE}

@@

@@

### How to turn timestep Z into annual Z for comparison?

Two approaches: sum nums each year, calculate survival between years (minus annual recruitment), turn into annual Z, or multiply timestep Zs to get annual?

### Biological sampling: true numbers at age class

@@

Everything below here is old.

We get true numbers at age at each output step by appling `sample_fish` to the survey_testNall result if the model group has one true age per output age class:

```{r truenatagecl, eval=FALSE}
# We need an effective N for the sample fish function, setting equal to actual N
# this one is high but not equal to total for numerous groups

# this doesnt work for sardines in CCA, change to full census

effNhigh <- data.frame(species=survspp, effN=rep(1e+8, length(survspp)))

numsallhigh <- sample_fish(survey_testNall, effNhigh)

#save for later use, takes a long time to generate
saveRDS(numsallhigh, file.path(d.name, paste0(scenario.name, "Natageclcensus.rds")))

@@

This is NOBA Cod true age class output. However, the 10 standard atlantis age classes aggregate true ages each for Cod.

@@

```{r Natage-plot, eval=FALSE}

@@

Natage_ss <- Natage[Natage$species == "North_atl_cod",]

@@

Perhaps overly optimistic, but if we feed these into `calc_stage2age` and get a result that matches the NOBA output ANNAGEBIO.nc, then we can assume the Zs are correct.

@@

```{r plotannage, eval=FALSE}

@@

Not looking great, but at least not constant. This could be wrong Zs or wrong calculations in `calc_stage2age`. Load true ages from ANNAGEBIO.nc:

```{r compareages, eval=FALSE}

@@

trueage_atout_ss <- trueage_atout[trueage_atout$species == "North_atl_cod",]

@@

```{r plot-esttrueage-atltrueage, eval=FALSE}

@@

We only have the first 10 cohorts of true age output due to hardcoding in load_nc. I may write another load function that changes that. Either way our early ages don't match. Something about the calculation or application of Z isn't correct.

>>>>

After several iterations of debugging, we are still not matching the pattern, but improved in that it isn't a constant punctuated by occasional high. Variation still doesn't look anything like the txt file either. The 5x yr timesteps are probably messing this up?

I've commented out the detailed debugging block below, but check out the rmd file if you are interested in the sausage making. Changes have been implemented in `calc_Z` so that it now apportions annual recruitment to timesteps and recruitment is correctly aligned with truth nums output.

However, it is not clear that subtracting off "recruitment" that has not been subject to mortality is appropriate when the output numbers at each timestep have been subject to mortality? I guess that was the input in each timestep as best we know it, but this is at best an approximation.
<!--
```{r calczsnippet, eval=FALSE}

#corrections listed here have been applied in calc_Z; this code not run

@@

### How to turn timestep Z into annual Z for proper comparison?

Two approaches: annual nums each year, calculate survival between years (minus annual recruitment), turn into annual Z, or add timestep Zs to get annual?

First an annual estimate based on end year numbers snapshots:

```{r annZ1}

totnums <- aggregate(atoutput ~ species + time, data = truenums_ss, sum) %>%
    mutate(time.days = (time+1)*runpar$toutinc) %>% #makes time 0 into days 0->73, etc
    mutate(yr = ceiling(time.days/365))  # yr 1 is 0:stepsperyr to match recruits yr1
    #mutate(yr = floor(time.days/365)) # includes 0 value, yr 1 is

#only want numbers at end of year, snapshot
totnumsann <- totnums %>%
  filter(time.days %in% seq(365, max(time.days), by=365))

# mg carbon converted to wet weight in tonnes
k_wetdry <- biol$kgw2d / 1000000000

# WARNING only works for CCA because YOY.txt rows 2:end are already in numbers
# and we are merging out the incorrect and irrelevant YOY row 1 (Time=0)
# also hardcoded for sardine example
recnums <- YOY_ss %>%
  mutate(yr = as.integer(round(YOY_ss$Time)/365)) %>%
  mutate(recnums = SAR.0/k_wetdry/biol$redfieldcn)

annZ1 <- merge(totnumsann, recnums) %>%
  mutate(recnums = replace_na(recnums, 0)) %>%
  mutate(numslessrec = atoutput - recnums) %>%
  mutate(surv = numslessrec/lag(atoutput, default = first(numslessrec))) %>%
  mutate(Z = -1 * log(surv))

plotZ <- ggplot() +
  geom_line(data=Zish, aes(x=Time/365, y=relZ)) +
  geom_point(data=annZ1, aes(x=Time/365, y=Z)) +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotZ + ylim(0, 2.0)

@@

This looks better; we are starting to see a pattern.

Second summing within year Z from calc_Z:

```{r annZ2}

annZ2 <- fullresZ %>%
  mutate(yr = floor(time/stepperyr)+1) %>%
  group_by(species, yr) %>%
  summarise(Z = sum(atoutput))

plotZ <- ggplot() +
  geom_line(data=Zish, aes(x=Time/365, y=relZ)) +
  geom_point(data=annZ2, aes(x=yr, y=Z)) +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotZ + ylim(0, 2.0)

Also looks better. Not sure which is more correct.

Compare methods:

```{r compZall}

plotZ <- ggplot() +
  geom_line(data=Zish, aes(x=Time/365, y=relZ, color="mort.txt Z")) +
  geom_point(data=annZ1, aes(x=Time/365, y=Z, color="endyr Z")) +
  geom_point(data=annZ2, aes(x=yr, y=Z, color="sum timestep Z")) +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotZ + ylim(0, 2.0)

Both seem to be tracking each other reasonably, and also general trends from mort.txt output. Discussion with Beth suggests that mort.txt output is an initial calculation combining deaths due to predation with deaths due to fishing, but that there is rescaling of predation (and possibly fishing?) afterward this output stage such that mort.txt does not provide final mortality. If the rescaling varies by timestep, then matching general patterns rather than the full internannual variability is probably the best we can do, since we don't know the full interannual variability.

We can probably conclude that the `calc_Z` function is working as correctly as possible now, given that it is an approximation no matter what, and the differences we see here are likley from temporal resolution of the Z estimate.

Let's see how it performs with another model.

### Test calc_Z NOBA

@@

```{r switch-NOBA}

initCCA <- FALSE
initNEUS <- FALSE
initNOBA <- TRUE

species_ss <- "North_atl_cod"

#Load functional groups
funct.groups <- load_fgs(dir=d.name,
                         file_fgs = functional.groups.file)
#Get just the names of active functional groups
funct.group.names <- funct.groups %>% 
  filter(IsTurnedOn == 1) %>%
  select(Name) %>%
  .$Name

#Get true NOBAaa
if(!file.exists(file.path(d.name, 
                          paste0("output", scenario.name, "run_truth.RData")))){
  #Store all loaded results into an R object
  truth <- run_truth(scenario = scenario.name,
                     dir = d.name,
                     file_fgs = functional.groups.file,
                     file_bgm = box.file,
                     select_groups = funct.group.names,
                     file_init = initial.conditions.file,
                     file_biolprm = biol.prm.file,
                     file_runprm = run.prm.file
  )
} else{
  truth <- get(load(file.path(d.name,
                              paste0("output", scenario.name, "run_truth.RData"))))
}

source(here("config/census_spec.R"))

@@

Lets try just the Z part for NOBA:

```{r Znoba}

# make a function for this
# add YOY file to the config files
YOY <- load_yoy(d.name, paste0("output", scenario.name, "YOY.txt"))

#oops, need to generalize calc_Z for subannual timesteps in YOY!
#or input YOY only at 0, 365, etc since the numbers repeat for timesteps 
YOY_ss <- YOY_ss %>%
  filter(Time %in% seq(0, max(Time), by=365))

# compare as above with mort.txt output
file.mort <- file.path(d.name, paste0("output", scenario.name, "Mort.txt"))

annZ2 <- fullresZ %>%
  mutate(yr = floor(time/stepperyr)+1) %>%
  group_by(species, yr) %>%
  summarise(Z = sum(atoutput))

plotZ <- ggplot() +
  geom_line(data=Zish, aes(x=Time/365, y=relZ, color="mort.txt Z")) +
  geom_point(data=fullresZ, aes(x=time/stepperyr, y=atoutput, color="each timestep Z")) +
  geom_point(data=annZ2, aes(x=yr-1, y=Z, color="sum timestep Z")) +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotZ + ylim(-1, 10.0)

This Z is a bad match. The Z values in NOBA's mort.txt are huge. Not sure what to make of this.

The above figure is now the best it is going to get after a deeper dive back into `calc_Z` for NOBA before we proceed to true ages. Found one bug hidden by sardine's equal recruit_period and recend, another where the recruitment period is all within one output timestep that resulted in proportions over 1 (inflating the recruitment and then subtracting it off). Further, it appears that timesteps for recruitment were still not correctly aligned, so trying a different alignment that leaves out the timestep 0 from the truth. Not sure the latter is correct, or if we need to rethink the calculation of survival based on when recruits actually enter the model.

Update: the new code actually makes the CCA sardine Z estimates more in line with each other, so the updates are correct and we are keeping them in `calc_Z`.

Most important: Beth says that the M in mort.txt can really be as far off as we are seeing it here. The green line in the above NOBA Z plot reflects the initial estimate of what dies by predation mortality, which is a gigantic overesimate because many availability parameters are set to 1. After this output step, the predation mortality is rescaled within the model to account for the fact that there really aren't that many to be eaten.

Our estimate of Z is from the actual numbers alive at the end of each timestep and the known input recruitment prior to any mortality, so I think it really is the best estimate we have. More debugging code commented out below but see the rmd file for the gory and redundant details.
<!--
```{r calcZ-snippet2, eval=FALSE}

# numbers at ageclaggregated over layers, retain polygons (output of create_survey)
survey_testN_ss <- survey_testNall[survey_testNall$species == species_ss,]

yoy <- YOY_ss
nums <- survey_testN_ss
fgs <- funct.groups
biolprm <- biol
toutinc <- runpar$toutinc

#calc_Z <- function(yoy, nums, fgs, biolprm, toutinc) {
  # subset the yoy for species included in the fgs file
  # that are turned on
  # colnames of the recruit data are "Time" and a column for each
  # species where the name is the functional group with an additional .0
  turnedon <- fgs[fgs$IsTurnedOn > 0, ]
  recruits <- yoy[, colnames(yoy) %in% c("Time", paste0(turnedon$Code, ".0"))]

  # merge recruits with strucn and resn of recruits from biol.prm file
  recruits <- merge(recruits, nitro[, c(1, 4)],
    by.x = "group", by.y = "1", all.x = TRUE, all.y = FALSE)
  colnames(recruits)[which(colnames(recruits) == "recruits")] <- "recruitsbio"
  # Get recruits in numbers rather than biomass
  recruits$recruits <- recruits$recruitsbio / recruits$sum
  recruits$yr <- as.integer(round(recruits$Time)/365) #needed to merge with totnums

  # June 2019:Isaac determined that yr 0 (Time 0) in the YOY file is not real, don't use it
  # code below assigns year 1 to times 0:stepperyr, all good output timesteps in truth$nums

  # June 12 2019 Beth determined that we do not need spawn_period

  # rectiming is a dataframe with species code, time_spawn (day of year),
  # recruit_time (number of days), and recruit period (number of days).
  # we use these to calculate recstart (day of year) and recend (day of year)
  rectiming <- merge(biolprm$time_spawn, biolprm$recruit_time, by = 1)
  rectiming <- merge(rectiming, biolprm$recruit_period, by = 1)
  names(rectiming) <- c("Code", "time_spawn", "recruit_time", "recruit_period")
  rectiming <- rectiming %>%
    mutate(recstart = time_spawn + recruit_time) %>%
    mutate(recend = recstart + recruit_period)

  # Sum numbers output over all boxes/depth/cohorts
  # align model output timesteps (days) with recruitment periods (days)
  totnums <- aggregate(atoutput ~ species + time, data = nums, sum) %>%
    #mutate(time.days = (time+1)*toutinc) %>% #makes time 0 into days 0->73, etc
    mutate(time.days = (time)*toutinc) %>% #makes time 1 into days 0->73, etc
    mutate(yr = ceiling(time.days/365))  # yr 1 is 0:stepsperyr to match recruits yr1

    recstart <- seq(recstart_temp$recstart[irow],by=365,length.out=nyrs)
    recstart <- recstart[recstart<max(totnums$Time[pick])]
    recend <- recstart + recstart_temp$recruit_period[irow]
    #rec_times <- rbind(rec_times,cbind(group,recstart,recend))

    for (i_rec in 1:length(recstart)) {
      i_tstart <- which(pick==min(pick[totnums$time.days[pick]>=recstart[i_rec]]))
      i_tstop <- which(pick==min(pick[totnums$time.days[pick]>=recend[i_rec]]))
      if (i_rec == length(recstart)) i_tstop <- length(pick)
      n_t <- 1+i_tstop-i_tstart
      for (i_t in 1:n_t) {
        t_temp <- totnums$time.days[pick[i_tstart+i_t-1]]
        num_temp <- t_temp - recstart[i_rec]
        if (i_t>1) {
          if ((recend[i_rec]-t_temp)>toutinc) {
            num_temp <- toutinc
          }
          else {
            num_temp <- recend[i_rec]-(totnums$time.days[pick[i_tstart+i_t-2]])
          }
        }
        frac_temp <- max(c(0,num_temp /(recend[i_rec]-recstart[i_rec])))
        if(frac_temp > 1.0) frac_temp = 1.0
        totnums$frac_recruit[pick[i_tstart+i_t-1]] <- frac_temp
      }
    }
  }

  totnums$group <- recruits$group[match(totnums$species, recruits$Name)]
  # For all time increments where there
  totnums$recruits[is.na(totnums$recruits)] <- 0

  totnums$annrecruits <- totnums$recruits

  #return(finaldata)
#}

## Biological sampling: true numbers at age class

@@

```{r Natage-plot}

@@

Natage_ss <- Natage[Natage$species == species_ss,]

@@

This really is still overly optimistic, but now that I think `calc_Z` is working, if we feed these into `calc_stage2age` and get a result that matches the NOBA output ANNAGEBIO.nc, then we can assume that function is correct.

@@

```{r plotannage}

# numbers at ageclaggregated over layers, retain polygons (output of create_survey)
survey_testN_ss <- survey_testNall[survey_testNall$species == species_ss,]

@@

Still not looking great, but at least not constant. I am now assuming we have the wrong calculations in `calc_stage2age`. Load true ages from ANNAGEBIO.nc:

```{r compareages}

@@

trueage_atout_ss <- trueage_atout[trueage_atout$species == species_ss,]

@@

```{r plot-esttrueage-atltrueage}

@@

We only have the first 10 cohorts of true age output due to hardcoding in load_nc. I may write another load function that changes that. Either way our early ages don't match. Need to step through `calc_stage2age` next!

### 2019-06-13 docs/index.Rmd
lets call calc_Z good

>>>>

June 7: `calc_Z` seems incorrect when compared to outputMort.txt; stepping through code needed

June 11: fixed matching of recruitment to timestep, still seems off

June 13: I think we have the best `calc_Z` we are going to get. After discussion with Beth, we should not expect to match output of Mort.txt because M gets rescaled so much. However, the comparison looks reasonable now for sardine and even for NOBA cod.

<a href="https://sgaichas.github.io/poseidon-dev/TestCalcZ2.html" target="_blank">Test Z calc and compare with output mort CCA and NOBA - June 7, 2019, updated June 13</a>
